Attribute fusion in a latent process model for time series of graphs

### 2011-12-16 ssp2011.tex
Resubmitted to IEEE TPAMI

\documentclass[10pt,journal,compsoc]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
%\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem Minh~Tang, Nam~H.~Lee
 and Carey~E.~Priebe are with the Department of Applied Mathematics
 and Statistics, Johns Hopkins University.\protect \\
  Email: \{mtang10, nhlee, cep\}@jhu.edu
\IEEEcompsocthanksitem Youngser Park is with the Center for Imaging
Science , Johns Hopkins University.\protect \\
  Email: youngser@jhu.edu}%
\thanks{}}
\markboth{IEEE Transactions on Pattern Analysis and Machine
  Intelligence (TPAMI)}{}
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
\begin{IEEEkeywords}
  Anomaly detection, Attributed Random Graphs, Random Dot
  Product Graphs.
\end{IEEEkeywords}}
\maketitle
\section{Introduction}

@@

@@

\section{Inference Example}
We present here an inference example with the following
parameters

@@

  \end{matrix}

@@

  \pi_0 = (0.10,0.30,0.60)^{T}, \quad \pi_1 = (0.25,0.40,0.35)^{T}

@@

The main implication that can
be inferred from \figurename~\ref{fig:power-scale} and
\figurename~\ref{fig:power-estimate}
is that inferential performance in the
mathematically tractable first-order and second-order approximation
models does provide guidance for methodological choices applicable to
the exact (realistic but intractable) model. Furthermore, to the
extent that the exact model is realistic, we may tentatively conclude
that approximation model investigations have some bearing on real
data applications.
\begin{figure*}[!t]
  \centering
  \subfloat[$\mathcal{E}_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-size.pdf}
  }
  \hfil
  \subfloat[$\Delta_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-maxd.pdf}
  }
  \hfil
  \subfloat[$\tau_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-scan.pdf}
  }
  \hfil
  \subfloat[$\Psi_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-ntri.pdf}
  }
  \caption{Power $\beta(\theta^{*}_r,r)$ at $\theta_{r}^{*} =
    \argmax_{\theta} \beta(\theta,r)$ as a function of $r$ for the
    four invariants at test size $\alpha = 0.05$. The horizontal lines
    represents first-order approximation $\pm$ three
    standard deviations, and the two curves represent the second
    approximation (green) and exact model (blue). The 10000 Monte
    Carlo replicates yields standard deviations not exceeding $0.005$ for the power
    estimates . The second approximation results match well with the exact model results, and
    both match well for large $r$ with the first-order approximation results.
  }
  \label{fig:power-scale}
\end{figure*}
\begin{figure*}[!t]
  \centering
  \subfloat[$\mathcal{E}_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-size.pdf}
  }
  \hfil
  \subfloat[$\Delta_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-maxd.pdf}
  }
  \hfil
  \subfloat[$\Psi_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scan.pdf}
    \label{fig5:subfig_scan}
  }
  \hfil
  \subfloat[$\tau_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-ntri.pdf}
  }
  \caption{Power $\beta$ as a function of angle $\theta$ for $\lambda
    = (cos(\theta),sin(\theta))$. The plot shows analytic asymptotic
    result (black), and first approximation (red), second
    approximation (green), and exact model (blue) estimates for the
    various invariants via Monte Carlo. The 10000 Monte Carlo
    replicates yields standard deviations not exceeding $0.005$ for
    the power estimates. The four vertical lines in each plot
    correspond to $\theta \in \{0, \pi/2, \pi/4,
    \theta^{*}\}$. Because we are considering a one-sided test and
    because $\pi_{1,k} > \pi_{0,k}$ for both $k = 1$ and $k = 2$, the
    power is maximized in the first quadrant, i.e., $\theta \in (0,
    \pi/2)$. The second approximation results match well
    with the exact model results, and both match well for large $r$
    with the first order approximation and the asymptotic results. The
    optimal $\theta^{*}$ is apparently different for the four different graph
    invariants.  
  }
  \label{fig:power-estimate}
  \end{figure*}
\section{Conclusions}
We have presented an analysis of change-point detection for time
series of attributed graphs through the use of test statistics which
are based on linear attribute fusion of some graph invariants. 
We derived the limiting distribution of these test statistics under
the assumptions that $n$, the number of vertices, is sufficiently
large and $r$, the process parameter rate, is also sufficiently
large. The limiting distribution for these test statistics are then
used to derive estimates for the power of the tests.

The experiment in \S \ref{eq:55} indicates that the power estimates
are accurate, even for the moderate value of $n = 100$. Furthermore,
it was also indicated that the optimal linear fusion parameter depends
on the graph invariant considered. In particular, the results depicted
in \figurename~\ref{fig:power-estimate} yield
$\theta^{*}_{\mathcal{E}_{\lambda}} \approx 0.24$,
$\theta^{*}_{\Delta_{\lambda}} \approx 0.32$,
$\theta^{*}_{\Psi_{\lambda}} \approx 0.14$ and
$\theta^{*}_{\tau_{\lambda}} \approx 0.38$. These optimal fusion
parameter differences are statistically significant, and combining
this result with the ``no uniformly most powerful invariant'' result
\cite{pao11:_statis_infer_random_graph}, we conclude that optimal
linear attribute fusion theory requires significant additional
development. Toward this end, the approximation models from
\cite{lee11} promise to be of assistance.

Hypothesis testing on time series of attributed graphs has
applications in diverse areas, e.g., social network analysis (wherein
vertices represent individual actors or organizations), connectome
inference (wherein vertices are neurons or brain regions) and text
processing (wherein vertices represent authors or documents). These
and many other applications may benefit from generalizations of our
results and the model in \cite{lee11} to directed, multi, and weighted
graphs, as well as the consideration of inference with errorful edge
attributes through an attribute confusion matrix.
% \appendix[Proofs of some stated results]
% \label{sec:proofs-some-stated}
% \begin{IEEEproof}[Corollary~\ref{cor:1}]
%   The maximizer of $\mu_\lambda$ also maximizes
%   \begin{equation*}
%     \mu_{\lambda}^{2} = \frac{\lambda^{T} \zeta \zeta^{T}
%       \lambda}{\lambda^{T} \xi \lambda}
%   \end{equation*}
%   Because $\xi$ is positive definite, there exists a positive definite matrix
%   $\xi^{1/2}$ such that $\xi^{1/2} \xi^{1/2} = \xi$. Letting $\nu = \xi^{1/2}
%   \lambda$, the above expression can be rewritten as
%   \begin{equation*}
%     \mu_{\lambda}^{2} = \frac{\nu^{T} \xi^{-1/2} \zeta \zeta^{T}
%       \xi^{-1/2} \nu}{ \nu^{T} \nu}
%   \end{equation*}
%   The claim then follows directly from the Rayleigh-Ritz theorem for
%   Hermitian matrices.
% \end{IEEEproof}
% \begin{IEEEproof}[Lemma~\ref{lem:3}]
%   $\tau_{\lambda}(G)$ is a U-statistics with kernel function
%   $h(Y_1, Y_2, Y_3) = Y_1 Y_2 Y_3$. By the theory of U-statistics, we
%   know that
%   \begin{equation}
%     \label{eq:48}
%     \frac{\tau_{\lambda}(G) -
%       \mathbb{E}[\tau_{\lambda}^{*}(G)]}{\sqrt{\mathrm{Var}[\tau_{\lambda}^{*}(G)]}}
%        \overset{\mathrm{d}}{\longrightarrow}  N(0,1)
%   \end{equation}
%   provided that $\mathrm{Var}[\tau_{\lambda}(G) -
%   \tau_{\lambda}^{*}(G)] = o(\mathrm{Var}[\tau_{\lambda}^{*}(G)])$.

>>>>

\documentclass[10pt,journal,compsoc]{IEEEtran}
\usepackage{fontenc}
\pdfminorversion=4
%\usepackage{lmodern}
\usepackage[utopia]{mathdesign}
\usepackage{graphicx}
%\usepackage{xltxtra}
%\setmainfont[Mapping=tex-text]{Linux Libertine}
\usepackage[cmex10]{amsmath}
\interdisplaylinepenalty=2500
\usepackage{array}
%\usepackage{amssymb}
\usepackage{amsthm}
%\usepackage[tight,normalsize,sf,SF]{subfigure}
\ifCLASSOPTIONcompsoc
\usepackage[font=normalsize,labelfont=sf,textfont=sf]{subfig}
\else
\usepackage[font=footnotesize]{subfig}
\fi
\usepackage{subfig}
\usepackage{parskip}
\usepackage{mathrsfs}
\usepackage{bm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\theoremstyle{definition}
\newtheorem{definition}{Definition}
\usepackage[colorlinks=true,pagebackref,linkcolor=magenta]{hyperref}
\usepackage[noadjust]{cite}
\renewcommand\arraystretch{1.2}
\newcommand{\argmax}{\operatornamewithlimits{argmax}}
\newcommand{\argmin}{\operatornamewithlimits{argmin}}
\def\mathllap{\mathpalette\mathllapinternal}
\def\mathllapinternal#1#2{%
\llap{$\mathsurround=0pt#1{#2}$}% $
}
\def\clap#1{\hbox to 0pt{\hss#1\hss}}
\def\mathclap{\mathpalette\mathclapinternal}
\def\mathclapinternal#1#2{%
\clap{$\mathsurround=0pt#1{#2}$}%
}
\def\mathrlap{\mathpalette\mathrlapinternal}
\def\mathrlapinternal#1#2{%
\rlap{$\mathsurround=0pt#1{#2}$}% $
}
\bibliographystyle{IEEEtran}
\begin{document}
\title{Attribute fusion in a latent process model for time series of
  graphs}
\author{Minh~Tang, Youngser~Park, Nam~H.~Lee, and Carey~E.~Priebe%
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem Minh~Tang, Nam~H.~Lee
 and Carey~E.~Priebe are with the Department of Applied Mathematics
 and Statistics, Johns Hopkins University.\protect \\
  Email: \{mtang10, nhlee, cep\}@jhu.edu
\IEEEcompsocthanksitem Youngser Park is with the Center for Imaging
Science , Johns Hopkins University.\protect \\
  Email: youngser@jhu.edu}%
\thanks{}}
\markboth{IEEE Transactions on Pattern Analysis and Machine
  Intelligence (TPAMI)}{}
\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
  Hypothesis testing on time series of attributed graphs has
  applications in diverse areas, e.g., social network analysis
  (wherein vertices represent individual actors or organizations),
  connectome inference (wherein vertices are neurons or brain regions)
  and text processing (wherein vertices represent authors or
  documents).  We consider the problem of anomaly/change point
  detection given the latent process model for time series of graphs
  with categorical attributes on the edges presented in \cite{lee11}.
  Various attributed graph invariants are considered, and their power
  for detection as a function of a linear fusion parameter is
  presented.  Our main result is that inferential performance in
  mathematically tractable first-order and second-order approximation
  models does provide guidance for methodological choices applicable
  to the exact (realistic but intractable) model. Furthermore, to the
  extent that the exact model is realistic, we may tentatively
  conclude that approximation model investigations have some bearing
  on real data applications.
\end{abstract}
\begin{IEEEkeywords}
  Anomaly detection, Attributed Random Graphs, Random Dot
  Product Graphs.
\end{IEEEkeywords}}
\maketitle
\section{Introduction}
\label{sec:introduction}

@@

\label{sec:latent-process-model}

@@

\section{Inference Examples}
We present here two inference examples, one on simulated data and the
other on the Enron email data set. In the first inference example, we
simulated data from the model in \S~\ref{sec:latent-process-model}
with the following parameters:

@@

  \end{matrix},

@@

  \pi_0 = (0.10,0.30,0.60)^{T}, \quad \pi_1 = (0.25,0.40,0.35)^{T}.

@@

The main implication that can
be inferred from \figurename~\ref{fig:power-scale} and
\figurename~\ref{fig:power-estimate}
is that inferential performance in the
mathematically tractable first-order and second-order approximation
models does provide guidance for methodological choices applicable to
the exact (realistic but intractable) model. Furthermore, to the
extent that the exact model is realistic, we may tentatively conclude
that approximation model investigations have some bearing on real
data applications.
\begin{figure*}[!t]
  \centering
  \subfloat[$\mathcal{E}_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-size.pdf}
  }
  \hfil
  \subfloat[$\Delta_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-maxd.pdf}
  }
  \hfil
  \subfloat[$\tau_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-scan.pdf}
  }
  \hfil
  \subfloat[$\Psi_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scale-ntri.pdf}
  }
  \caption{Power $\beta(\theta^{*}_r,r)$ at $\theta_{r}^{*} =
    \argmax_{\theta} \beta(\theta,r)$ as a function of $r$ for the
    four invariants at test size $\alpha = 0.05$. The horizontal lines
    represents first-order approximation $\pm$ three
    standard deviations, and the two curves represent the second
    approximation (green) and exact model (blue). The 10000 Monte
    Carlo replicates yields standard deviations not exceeding $0.005$ for the power
    estimates . The second approximation results match well with the exact model results, and
    both match well for large $r$ with the first-order approximation results.
  }
  \label{fig:power-scale}
\end{figure*}
\begin{figure*}[!t]
  \centering
  \subfloat[$\mathcal{E}_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-size.pdf}
  }
  \hfil
  \subfloat[$\Delta_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-maxd.pdf}
  }
  \hfil
  \subfloat[$\Psi_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-scan.pdf}
    \label{fig5:subfig_scan}
  }
  \hfil
  \subfloat[$\tau_{\lambda}$]{
    \includegraphics[width=7cm]{graphics/pwr-ntri.pdf}
  }
  \caption{Power $\beta$ as a function of angle $\theta$ for $\lambda
    = (cos(\theta),sin(\theta))$. The plot shows analytic asymptotic
    result (black), and first approximation (red), second
    approximation (green), and exact model (blue) estimates for the
    various invariants via Monte Carlo. The 10000 Monte Carlo
    replicates yields standard deviations not exceeding $0.005$ for
    the power estimates. The four vertical lines in each plot
    correspond to $\theta \in \{0, \pi/2, \pi/4,
    \theta^{*}\}$. Because we are considering a one-sided test and
    because $\pi_{1,k} > \pi_{0,k}$ for both $k = 1$ and $k = 2$, the
    power is maximized in the first quadrant, i.e., $\theta \in (0,
    \pi/2)$. The second approximation results match well
    with the exact model results, and both match well for large $r$
    with the first order approximation and the asymptotic results. The
    optimal $\theta^{*}$ is apparently different for the four different graph
    invariants.  
  }
  \label{fig:power-estimate}
\end{figure*}

Our second example uses the Enron data set which consists of email
messages between 184 employees of the Enron corporation during a time
period from 1998 to 2002. The Enron email messages for most of the
2001 calendar year have been annotated into 32 different topics 
\cite{berry01:_topic_annot_enron_email_data_set} and thus form the
basis of our analysis. We chose to condense these 32 different
topics into two groups. The first group consists of
general enery-related topics and the second group consists of
topics that are specifically related to the Enron corporation. 
From the collection of email messages, we construct a time series of
graphs. Each graph consists of email messages sent during a week in
2001. The vertices of the graphs represent the employees and the edges
represent email communication between the employees. The attributes on
the edges are given by the grouping of the email topics as mentioned
above.

The detection of a chatter anomaly in the Enron data set during the
first week of May 2001 was previously
reported in \cite{priebe05:_scan_statis_enron_graph}, but with 
unattributed edges. We chose to investigate the effect
of attribute fusion for that same week in our inference example. 
\figurename~\ref{fig:enron} presents a plot of the normalized
statistic $T_{\lambda}^{l}$ with $l = 20$ for the four graph
invariants against the fusion parameter
$\lambda = (cos(\theta), sin(\theta))$. \figurename~\ref{fig:enron}
indicates that the fusing of attributes leads to better detection
as exemplified by the fact that the maximum for each of the normalized
statistics $T_{\lambda}$ occurs for $\theta \not \in
\{0,\pi/2\}$. Furthermore, the optimal fusion parameters are dependent
on the specific graph invariants.
\section{Conclusions}
We have presented an analysis of change-point detection for time
series of attributed graphs through the use of test statistics which
are based on linear attribute fusion of some graph invariants. 
We derived the limiting distribution of these test statistics under
the assumptions that $n$, the number of vertices, is sufficiently
large and $r$, the process parameter rate, is also sufficiently
large. The limiting distribution for these test statistics are then
used to derive estimates for the power of the tests.

The simulation experiment in \S \ref{eq:55} indicates that the power estimates
are accurate, even for the moderate value of $n = 100$. Furthermore,
it was also indicated that the optimal linear fusion parameter depends
on the graph invariant considered. In particular, the results depicted
in \figurename~\ref{fig:power-estimate} yield
$\theta^{*}_{\mathcal{E}_{\lambda}} \approx 0.24$,
$\theta^{*}_{\Delta_{\lambda}} \approx 0.32$,
$\theta^{*}_{\Psi_{\lambda}} \approx 0.14$ and
$\theta^{*}_{\tau_{\lambda}} \approx 0.38$. These optimal fusion
parameter differences are statistically significant, and combining
this result with the ``no uniformly most powerful invariant'' result
\cite{pao11:_statis_infer_random_graph}, we conclude that optimal
linear attribute fusion theory requires significant additional
development. Toward this end, the approximation models from
\cite{lee11} promise to be of assistance.

Hypothesis testing on time series of attributed graphs has
applications in diverse areas, e.g., social network analysis (wherein
vertices represent individual actors or organizations), connectome
inference (wherein vertices are neurons or brain regions) and text
processing (wherein vertices represent authors or documents). These
and many other applications may benefit from generalizations of our
results and the model in \cite{lee11} to directed, multi, and weighted
graphs, as well as the consideration of inference with errorful edge
attributes through an attribute confusion matrix.
\begin{figure}[tbp]
  \centering
 \includegraphics[width=8cm]{graphics/enron152.pdf} 
  \caption{Detecting chatter anomaly in the Enron time series of
    graphs. The graph in question corresponds to the first week of May
    2001. Attribute fusion provides superior detection.}
  \label{fig:enron}
\end{figure}
% \appendix[Proofs of some stated results]
% \label{sec:proofs-some-stated}
% \begin{IEEEproof}[Corollary~\ref{cor:1}]
%   The maximizer of $\mu_\lambda$ also maximizes
%   \begin{equation*}
%     \mu_{\lambda}^{2} = \frac{\lambda^{T} \zeta \zeta^{T}
%       \lambda}{\lambda^{T} \xi \lambda}
%   \end{equation*}
%   Because $\xi$ is positive definite, there exists a positive definite matrix
%   $\xi^{1/2}$ such that $\xi^{1/2} \xi^{1/2} = \xi$. Letting $\nu = \xi^{1/2}
%   \lambda$, the above expression can be rewritten as
%   \begin{equation*}
%     \mu_{\lambda}^{2} = \frac{\nu^{T} \xi^{-1/2} \zeta \zeta^{T}
%       \xi^{-1/2} \nu}{ \nu^{T} \nu}
%   \end{equation*}
%   The claim then follows directly from the Rayleigh-Ritz theorem for
%   Hermitian matrices.
% \end{IEEEproof}
% \begin{IEEEproof}[Lemma~\ref{lem:3}]
%   $\tau_{\lambda}(G)$ is a U-statistics with kernel function
%   $h(Y_1, Y_2, Y_3) = Y_1 Y_2 Y_3$. By the theory of U-statistics, we
%   know that
%   \begin{equation}
%     \label{eq:48}
%     \frac{\tau_{\lambda}(G) -
%       \mathbb{E}[\tau_{\lambda}^{*}(G)]}{\sqrt{\mathrm{Var}[\tau_{\lambda}^{*}(G)]}}
%        \overset{\mathrm{d}}{\longrightarrow}  N(0,1)
%   \end{equation}
%   provided that $\mathrm{Var}[\tau_{\lambda}(G) -
%   \tau_{\lambda}^{*}(G)] = o(\mathrm{Var}[\tau_{\lambda}^{*}(G)])$.
